Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-10-22 _tex/index.tex
Built site for gh-pages

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={causal inference, probability, Thurstone, comparative
judgement, directed acyclic graph, structural causal models, statistical
modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Causes and effects in Dichotomous Comparative Judgments: an
information-theoretical system of plausible mechanism}
\author[1]{Jose Manuel Rivera Espejo%
\corref{cor1}%
}
 \ead{JoseManuel.RiveraEspejo@uantwerpen.be} 
\author[1]{Tine van Daal%
%
}
 \ead{tine.vandaal@uantwerpen.be} 
\author[1]{Sven De Maeyer%
%
}
 \ead{sven.demaeyer@uantwerpen.be} 
\author[2]{Steven Gillis%
%
}
 \ead{steven.gillis@uantwerpen.be}

@@

Over the past decade, numerous studies have documented the effectiveness
of the \emph{comparative judgment} (CJ) method \citep{Thurstone_1927}
for assessing competencies and traits. These studies have evaluated CJ
from two main perspectives: its ability to produce reliable and valid
trait scores, and its practical applicability. In terms of reliability
and validity, research has shown that CJ can generate precise and
consistent scores that accurately represent the traits being measured.
Notable contributions in this research area include studies by Pollit
\citetext{\citeyear{Pollitt_2012a}; \citeyear{Pollitt_2012b}},
\citet{Whitehouse_2012}, \citet{vanDaal_et_al_2016},
\citet{Lesterhuis_2018}, \citet{Bramley_et_al_2019},
\citet{Verhavert_et_al_2019}, \citet{Crompvoets_et_al_2022}, and
\citet{Bouwer_et_al_2023}. Regarding practical applicability, several
studies have highlighted the method's versatility in both educational
and non-educational contexts, presenting it as an efficient and
effective alternative for measurement and evaluation. Key examples in
this research area include the works of \citet{Jones_2015},
\citet{Bartholomew_et_al_2018}, \citet{Jones_et_al_2019},
\citet{Marshall_et_al_2020}, \citet{Bartholomew_et_al_2020}, and
\citet{Boonen_et_al_2020}.

Despite the growing number of CJ studies, the unsystematic and
fragmented research approaches employed in the literature have
overlooked several critical issues concerning CJ. These issues can be
categorized into concerns about the method's structural model,
measurement model, and experimental design.

A key issue in the first category is the apparent disconnect between the
method's structural and measurement models. A common practice in CJ
literature involves performing data analysis and hypothesis testing on
scores previously estimated using the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. These studies use the scores or
their transformations to identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2017}, detect `bias' in judges'
ratings \citep{Pollitt_et_al_2003, Pollitt_2012b}, or test various
hypothesis
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
However, since CJ scores are parameter estimates with inherent
uncertainty, the statistical literature suggests that separating the
analysis from this uncertainty may artificially inflate the precision
and power of the results. This, in turn, could increase the risk of
committing a type I error, where a null hypothesis is wrongly rejected
\citep{McElreath_2020}.

@@

\subsection{Let's talk about Thurstone co.}\label{sec-theory-thurstone}

\subsection{A scientific model for the CJ}\label{sec-theory-scientific}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={causal inference, probability, Thurstone, comparative
judgement, directed acyclic graph, structural causal models, statistical
modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Let's talk about Thurstone \& Co.: An information-theoretical
model for comparative judgments, and its statistical translation}
\author[1]{Jose Manuel Rivera Espejo%
\corref{cor1}%
}
 \ead{JoseManuel.RiveraEspejo@uantwerpen.be} 
\author[1]{Tine van Daal%
%
}
 \ead{tine.vandaal@uantwerpen.be} 
\author[1]{Sven De Maeyer%
%
}
 \ead{sven.demaeyer@uantwerpen.be} 
\author[2]{Steven Gillis%
%
}
 \ead{steven.gillis@uantwerpen.be}

@@

Over the past decade, numerous studies have documented the effectiveness
of the \emph{comparative judgment} (CJ) method \citep{Thurstone_1927}
for assessing competencies and traits. These studies have evaluated CJ
from three main perspectives: its ability to produce reliable and valid
trait scores, its practical applicability, and its time efficiency.
Research on reliability and validity shows that CJ can generate precise
and consistent scores
\citep{Pollitt_2012a, Pollitt_2012b, Coertjens_et_al_2017, Goossens_et_al_2018, Verhavert_et_al_2019, Crompvoets_et_al_2022, Bouwer_et_al_2023}
that accurately represent the traits being measured
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bouwer_et_al_2023}.
Research on practical applicability highlights CJ's versatility across
both educational and non-educational contexts, presenting it as an
efficient and effective alternative for measurement and evaluation
\citep{Pollitt_2004, Jones_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.
Lastly, research on time efficiency suggests that CJ can offer at least
equal, if not significant, time savings when evaluating stimuli compared
to traditional marking methods
\citep{Pollitt_2012a, Pollitt_2012b, Coertjens_et_al_2017, Goossens_et_al_2018}.

Nevertheless, despite the growing number of CJ studies, the unsystematic
and fragmented research approaches employed in the literature have
overlooked several critical issues concerning the method. These issues
fall into three main categories: the disconnect between CJ's structural
and measurement model, the over-reliance on Thurstone's Case 5
\citeyearpar{Thurstone_1927} in its measurement model, and the role of
the comparison algorithms and the number of comparisons on the method's
reliability and validity. In the following sections, each issue will be
discussed in detail, followed by the introduction of a theoretical model
and its translation into a statistical model that addresses all three
concerns simultaneously.

\section{Overlooked issues in CJ literature}\label{sec-theory-issues}

\subsection{The disconnect between structural and measurement
model}\label{the-disconnect-between-structural-and-measurement-model}

In the CJ literature, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model
for CJ because it specifies how latent variables are estimated from
observed ones \citep{Hoyle_et_al_2023, Kline_et_al_2023}. In a CJ study,
multiple judges conduct several rounds of pairwise comparisons to
evaluate the relative manifestation of a trait between a pair of
stimuli. This evaluation results in a dichotomous outcome that indicates
which stimulus is perceived to exhibit a higher degree of the trait. The
BTL model then uses these observed outcomes to estimate scores that
represent the latent trait of interest
\citep{Pollitt_2012a, Pollitt_2012b, Whitehouse_2012, Jones_2015, vanDaal_et_al_2016, Lesterhuis_2018, Boonen_et_al_2020, Bouwer_et_al_2023}.

Moreover, in the CJ literature, it is common practice for BTL-generated
scores, or their transformations, to undergo additional data analysis
and hypothesis testing. Researchers have utilized these scores in
independent analyses to identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other scoring methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

\subsection{Case 5 and the measurement
model}\label{case-5-and-the-measurement-model}

\subsection{The efficiency and reliability of comparison
algorithms}\label{the-efficiency-and-reliability-of-comparison-algorithms}

\subsection{What is the solution}\label{what-is-the-solution}

@@

\subsection{A theoretical model for the CJ}\label{sec-theory-scientific}
